The Impact of Aging Population and Education Attainment on Fertility Rate in US over 1990 to 2020

### 2024-04-21 paper/paper.qmd
Resolve comments from Rohan

---
title: "The Impact of Aging Population and Education Attainment on Fertility Rate in US over 1990 to 2020"
author: 
  - Wanling Ma
thanks: "Code and data are available at: https://github.com/WanlingMa/The-Impact-of-Aging-Population-and-Education-Attainment-on-Fertility-Rate-in-US-over-1990-to-2020"
date: today
date-format: long
abstract: "This paper investigates the relationship between the aging population rate, educational attainment, and fertility rates in the United States over a thirty-year period from 1990 to 2020. Through a detailed analysis using data from the United Nations and the World Bank, we found that higher educational attainment is associated with lower fertility rates, while an increase in the working-age population percentage positively affects fertility. These findings highlight the significant impact of demographic and educational shifts on the nation's fertility trends. Understanding these dynamics is crucial for developing policies that address the challenges and opportunities of an evolving society."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

The findings of this study are revealing. Analysis through a linear regression model indicates a distinct inverse relationship between the fertility rate and educational attainment: as the average level of education rises, fertility rates tend to decline. This trend suggests that higher education may contribute to delayed family planning and a reduction in the number of children per family. Additionally, this paper reveals that the increase in the working-age population percentage is positively associated with higher fertility rates. This correlation implies that, conversely, an aging population tends to have a negative impact on fertility rates.

@@

The paper is organized into sections covering [Data], [Model], [Results], and [Discussions]. The [Data] section outlines our utilization of datasets from the United Nations and the World Bank, including our methodologies for data selection, cleaning, and the challenges encountered. The [Model] section introduces our linear regression analysis, detailing the construction and justification of the model used to examine the impact of the aging population rate and educational attainment on fertility rates. In the [Results] section, we disclose our principal discoveries, highlighting the positive impact of the working-age population percentage on fertility rates and the inverse relationship between educational attainment and fertility. Finally, the [Discussions] section melds our findings with the existing body of research, interpreting the implications of these trends and suggesting avenues for future investigations that further explore these complex relationships.

## Estimand

@@

## Data Clean and Processing

@@

After @OurWorldInData's processing, we further processed the data using R programming [@citeR]. This involved reading raw data files into data frames, filtering the datasets to include only the United States from 1990 to 2020 using functions from `dplyr` packages [@citeDplyr]. Additionally, we leveraged the ggplot2 package to create visual representations of these trends, such as stacked area charts for population distribution by age groups, line graphs for the fertility rate, and a line chart for tertiary education enrollment rates [@citeGgplot2].

@@

The datasets were collected and measured using authoritative sources, primarily the United Nations and the World Bank. The United Nations Department of Economic and Social Affairs Population Division offers detailed demographic data, including age distribution and fertility rates, through regular population surveys and censuses conducted in member countries. Similarly, the World Bank compiles education and economic indicators, such as tertiary education enrollment rates, from official national and international statistical sources. This rigorous data collection and measurement process involves data cleaning, validation, and standardization efforts to facilitate global comparability and to address discrepancies in data reporting standards across countries.

@@

The education achievements dataset, visualized by the @fig-education-levels, presents the tertiary education enrollment rate in the United States from 1990 to 2020. Notably, the data reveals a general upward trend in enrollment rates, starting from 69.67681% in 1990 and peaking at 88.88941% in 2015, before slightly declining to 87.56766% by 2020. The dataset highlights significant growth in tertiary education participation, reflecting an increasing value placed on higher education in the U.S. over the period analyzed.

There is a noticeable dip in enrollment rate to 70.50774% in 1998, following a period of steady increases. This decline is an anomaly within an otherwise consistent upward trajectory. Following this dip, the rates resumed their increase, suggesting a temporary setback rather than a long-term trend. The data underscores the fluctuating yet overall positive trajectory of tertiary education enrollment in the U.S., emphasizing the nation's emphasis on higher education as a key component of societal progress.

@@

From @fig-fertility-rate, we can see that this dataset charts the fertility rate trend in the United States from 1990 to 2020, showing the average number of children born per woman during this period. The initial rate in 1990 stood at 2.0692 children per woman, experiencing slight fluctuations through the decades, peaking at 2.1118 in 2007, and then demonstrating a notable decline to 1.6428 by 2020. This progression indicates a significant shift in family planning and societal trends, with the fertility rate decreasing below the replacement level of approximately 2.1 children per woman, which is generally considered necessary to maintain a stable population level.

>>>>

---
title: "The Impact of Aging Population and Education Attainment on Fertility Rate in US over 1990 to 2020"
author: 
  - Wanling Ma
thanks: "Code and data are available at: https://github.com/WanlingMa/The-Impact-of-Aging-Population-and-Education-Attainment-on-Fertility-Rate-in-US-over-1990-to-2020"
date: today
date-format: long
abstract: "This paper investigates the relationship between the aging population rate, educational attainment, and fertility rates in the United States over a thirty-year period from 1990 to 2020. Data from the United Nations and the World Bank shows that higher educational attainment is associated with lower fertility rates, while an increase in the working-age population percentage positively affects fertility. These findings highlight the impact of demographic and educational shifts on the nation's fertility trends."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

We find a distinct inverse relationship between the fertility rate and educational attainment: as the average level of education rises, fertility rates tend to decline. This trend suggests that higher education may contribute to delayed family planning and a reduction in the number of children per family. Additionally, this paper reveals that the increase in the working-age population percentage is positively associated with higher fertility rates. This correlation implies that, conversely, an aging population tends to have a negative impact on fertility rates.

@@

The paper is organized into sections covering [Data], [Model], [Results], and [Discussion]. The [Data] section outlines our datasets, including data selection, cleaning, and the challenges encountered. The [Model] section introduces our linear regression analysis, detailing the construction and justification of the model used to examine the impact of the aging population rate and educational attainment on fertility rates. In the [Results] section, we highlight our finding of the positive impact of the working-age population percentage on fertility rates and the inverse relationship between educational attainment and fertility. Finally, the [Discussion] section considers the existing body of research, interpreting the implications of these trends and suggesting avenues for future investigations that further explore these complex relationships.

@@

## Data Cleaning and Processing

@@

After @OurWorldInData's processing, we further processed the data using R programming [@citeR]. This involved reading raw data files into data frames, filtering the datasets to include only the United States from 1990 to 2020 using functions from `dplyr` packages [@citeDplyr]. Additionally, we used ggplot2 to create visual representations of these trends [@citeGgplot2].

@@

The datasets were collected and measured using authoritative sources, primarily the United Nations and the World Bank. The United Nations Department of Economic and Social Affairs Population Division offers detailed demographic data, including age distribution and fertility rates, through regular population surveys and censuses conducted in member countries. Similarly, the World Bank compiles education and economic indicators, such as tertiary education enrollment rates, from official national and international statistical sources. This data collection and measurement process involves data cleaning, validation, and standardization efforts to facilitate global comparability and to address discrepancies in data reporting standards across countries.

@@

The education achievements dataset, visualized by the @fig-education-levels, presents the tertiary education enrollment rate in the United States from 1990 to 2020. Notably, the data reveals a general upward trend in enrollment rates, starting from 70% in 1990 and peaking at 89% in 2015, before slightly declining to 88% by 2020. The dataset highlights significant growth in tertiary education participation, reflecting an increasing value placed on higher education in the U.S. over the period analyzed.

There is a noticeable dip in enrollment rate to 71% in 1998, following a period of steady increases. This decline is an anomaly within an otherwise consistent upward trajectory. Following this dip, the rates resumed their increase, suggesting a temporary setback rather than a long-term trend. The data underscores the fluctuating yet overall positive trajectory of tertiary education enrollment in the U.S., emphasizing the nation's emphasis on higher education as a key component of societal progress.

@@

From @fig-fertility-rate, we can see that this dataset charts the fertility rate trend in the United States from 1990 to 2020, showing the average number of children born per woman during this period. The initial rate in 1990 stood at 2.07 children per woman, experiencing slight fluctuations through the decades, peaking at 2.10 in 2007, and then demonstrating a notable decline to 1.60 by 2020. This progression indicates a significant shift in family planning and societal trends, with the fertility rate decreasing below the replacement level of approximately 2.10 children per woman, which is generally considered necessary to maintain a stable population level.
